Iterative Monomorphisation

### 2025-07-02 iterative_monomorphisation.tex
did

One approach \cite{mono-trans} is to encode a complete polymorphic type system, %using dedicated function or predicate symbols in a monomorphic logic,
but this increases the size of the input problem substantially and slows down provers \cite{mono-trans}.
%
Another approach to encode polymorphism is based on {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}.
%
Iterative monomorphisation heuristically instantiates the formulae's type variables with concrete types.
%
However, any translation relying on a finite number of instantiations is inevitably incomplete.
By a typed version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of~\(\varphi\), but
it cannot be computed in general~\cite[Theorem~1]{expr-poly-types}.

B\"ohme's approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}. This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers. However, it is documented only as a single subsection in B\"ohme's PhD thesis~\cite[Section 2.2.1]{sb-phd}.
\relax{SMT-LIB} also uses iterative monomorphisation \cite{smt-lib-mono}.
Moreover, a similar algorithm appears to be implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.

>>>>

\looseness=-1
To achieve this, one approach \cite{mono-trans} is to encode a complete polymorphic type system, %using dedicated function or predicate symbols in a monomorphic logic,
but this increases the size of the input problem substantially and slows down provers \cite{mono-trans}.
%
Another approach to encode polymorphism is based on {iterative monomorphisation}, as described by B\"ohme~\cite[Section 2.2.1]{sb-phd}.
%
Iterative monomorphisation heuristically instantiates the formulae's type variables with concrete types.
%
However, any translation relying on a finite number of instantiations is inevitably incomplete.
By a typed version of the compactness theorem, for any first order polymorphic formula \(\varphi\), there exists an equisatisfiable finite set of monomorphic instances of~\(\varphi\), but
it cannot be computed in general~\cite[Theorem~1]{expr-poly-types}.

B\"ohme's approach is implemented as part of Isabelle/HOL's SMT (satisfiability modulo theories) integration~\cite[Chapter 2]{sb-phd}. This implementation is also used by Sledgehammer~\cite{judgement,paulson-blanchette-2010} to interface with superposition based automatic theorem provers. However, it is documented only superficially~\cite[Section 2.2.1]{sb-phd}.
An iterative monomorphisation approach is also described in the context of SMT-LIB \cite{smt-lib-mono}.
Moreover, a similar algorithm appears to have been implemented in the \relax{MESON} tactic~\cite{harrison-1996} of HOL Light, but it is undocumented.
